\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2024-06-10 doc/paper-jss/sdmTMB-paper.Rnw
Edits

x <- seq(0, 1, length.out = 200)
r <- seq(0.2, 1, 0.2)
r <- c(0.2, 0.6)
df <- data.frame(
  x = rep(x, length(r)),
  range = rep(r, each = length(x))
)
matern <- function(h, sigma = 1, kappa, nu = 1) {
  ret <- (sigma^2/(2^(nu - 1) * gamma(nu))) *
    ((kappa * abs(h))^nu) *
    besselK(kappa * abs(h), nu)
  ret
  ret[x == 0] <- sigma^2
  ret
}
blues <- RColorBrewer::brewer.pal(length(r) + 1, "Blues")[-1]
df$cor <- matern(df$x, kappa = sqrt(8) / df$range)
sim_g3 <- ggplot(df, aes(x, cor, col = as.factor(range), group = as.factor(range))) +
  geom_line() +
  theme_light() +
  xlab("Distance") +
  ylab("Correlation") +
  labs(colour = "Range") +
  coord_cartesian(expand = FALSE, ylim = c(0, 1)) +
  scale_colour_manual(values = blues) +
  geom_hline(yintercept = 0.13, col = "grey50", lty = 2) +
  scale_x_continuous(breaks = r) +
  theme(legend.position = c(0.7, 0.7)) +
  ggtitle("(c) Matérn correlation function")

@@

% We omit example code for Moran's I calculation for brevity but include it in the code supplement.
\added{
We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but a lack of evidence for spatial correlation after including a random field (p $>$ 0.01) suggesting that residual spatial autocorrelation is alleviated by including the random field.
The specific p-value is dependent on the seed due to the randomization in the randomized quantile residuals.
}

@@

\added{
where $n$ represents the number of left-out data points, $\log \mathcal{L}$ represents the log likelihood, $y_i$ represents left-out data point $i$, and $\widehat{y_i}$ represents the prediction for left-out data point $i$ with the fixed effects at their MLEs and random effects at their empirical Bayes estimates.
}

@@

In practice, we would \deleted{want to} repeat this procedure several times to ensure the rank order is \added{insensitive} to the randomly chosen folds and, if it was, consider averaging across multiple folds or increasing the number of folds.

@@

<<dog-index, echo=TRUE>>=
pred2 <- predict(fit, newdata = grid, return_tmb_object = TRUE)
ind <- get_index(pred2, bias_correct = TRUE, area = rep(4, nrow(grid)))
@

@@

In our final example, we demonstrate a model with spatially varying coefficient (SVC) effects and illustrate combining uncertainty from parameters by working with draws from the joint parameter precision matrix.
SVC models are a class of models in which coefficients are allowed to vary spatially constrained by some smooth function \citep{hastie1993, thorson2023SVC}.
In addition, we illustrate using restricted maximum likelihood (REML), which integrates over both the fixed and random effects.

@@

snow_g1 <- ggplot(data = nao_effect_df) +
  geom_sf(data = land, fill = "white", colour = "white", lwd = 0.35) +
  geom_sf(data = lakes, colour = "gray23", fill = "grey90", lwd = 0.35) +
  geom_point(aes(X, Y, colour = nao_effect, size = count), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50", fill = NA, lwd = 0.35) +
  geom_sf(data = lakes, colour = "gray50", fill = NA, lwd = 0.35) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr)),
    guide = guide_colourbar(direction = "horizontal", title.vjust = 1, title.position = "top", label.position = "bottom")
  ) +
  guides(size = "none") +
  ggtitle(paste0("(a) Median multiplicative effect: ", round(min(snow$nao_effect), 2), " to ", round(max(snow$nao_effect), 2))) +
  labs(colour = "NAO effect\non Snowy Owl counts") +
  theme_bw() +
  theme(
    legend.position = c(0.25, 0.17),
    legend.title = element_text(size = 9, hjust = 0),
    plot.margin = unit(c(0, 0.5, 0, 0), "cm"),
    panel.background = element_rect(fill = "grey90", colour = NA),
    panel.grid.major = element_line(colour = "white", linewidth = 0.5),
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.box.background = element_rect(fill = "transparent", colour = NA),
    plot.title = element_text(size = 10, hjust = -0.15),
    axis.title = element_blank()
  )

th <- theme(
    legend.position = "none",
    panel.background = element_rect(fill = "grey90", colour = NA),
    panel.grid.major = element_line(colour = "white", size = 0.5),
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.box.background = element_rect(fill = "transparent", colour = NA),
    plot.title = element_text(size = 10), axis.ticks = element_blank(),
    axis.title = element_blank(), axis.text = element_blank()
  )

snow_g2 <- ggplot(data = nao_effect_df) +
  geom_sf(data = land, fill = "white", colour = "white", lwd = 0.35) +
  geom_sf(data = lakes, colour = "gray23", fill = "grey90", lwd = 0.35) +
  geom_point(aes(X, Y, colour = nao_effect_lwr, size = count), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50", lwd = 0.35) +
  geom_sf(data = lakes, colour = "gray50", fill = NA, lwd = 0.35) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr))
  ) +
  ggtitle(paste0("(b) Lower 80% CI: ", round(min(snow$nao_effect_lwr), 2), " to ", round(max(snow$nao_effect_lwr), 2))) +
  theme_bw() +
  th

snow_g3 <- ggplot(data = nao_effect_df) +
  geom_sf(data = land, fill = "white", colour = "white", lwd = 0.35) +
  geom_sf(data = lakes, colour = "gray23", fill = "grey90", lwd = 0.35) +
  geom_point(aes(X, Y, colour = nao_effect_upr, size = count), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50", lwd = 0.35) +
  geom_sf(data = lakes, colour = "gray50", fill = NA, lwd = 0.35) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr))
  ) +
  ggtitle(paste0("(c) Upper 80% CI: ", round(min(snow$nao_effect_upr), 2), " to ", round(max(snow$nao_effect_upr), 2))) +
  theme_bw() +
  th

@@

\begin{figure}[htb]
\centering
% \includegraphics[width=3.5in]{../../figs/timing-spatial-2023-03-16-xkcd.pdf}
\includegraphics[width=\textwidth]{figs/timing-spatial-multipanel.pdf}
\caption[]{
Comparison of time to fit an SPDE spatial random field model with an intercept, one fixed-effect predictor, Gaussian error, and a sequence of SPDE resolutions to \added{three dataset sizes}.
Lines represent means and ribbons 90\% quantiles across 50 random iterations.
\pkg{VAST} and \pkg{tinyVAST} should be similar to \sdmTMB\ and so are not shown.
\pkg{inlabru} \added{(version 2.10.1)} used the empirical Bayes integration strategy and Gaussian approximation with \code{bru\_max\_iter = 1}, and the \fct{like} formulation.
\pkg{mgcv} \added{(version 1.9.1)} used \fct{bam}, \code{method = `fREML'}, and discretized covariates \citep{miller2019}.
Note that \pkg{spaMM} \added{(version 4.4.16)} only fits spatial, not spatiotemporal, models.
All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.
\added{See Figure~\ref{fig:timing2} for a version with log-log axes.}
}
\label{fig:timing}
\end{figure}

@@

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table~\ref{tab:functionality}), we suggest the primary differences are the user interface and speed.
We think users familiar with \fct{stats::glm}, \pkg{lme4}, \pkg{mgcv}, or \pkg{glmmTMB} will find \sdmTMB\ most approachable.
Users familiar with \INLA\ will find \pkg{inlabru} approachable.
Users familiar with \pkg{mgcv} can adapt \pkg{mgcv} to fit similar models with custom code \citep{miller2019} and \INLA/\pkg{inlabru} and \pkg{mgcv} are also general purpose modelling packages.
\pkg{VAST} and \pkg{tinyVAST} are the sole options for fitting some multivariate models; alternatively, because these packages focus on multivariate delta models and fisheries applications, users fitting ``simple'' univariate spatial/spatiotemporal GLMMs will likely find \sdmTMB\ more straightforward.
Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \sdmTMB\ or \pkg{VAST} (although such quantities can be calculated using other packages post hoc).
Overall, we think there is value in that \sdmTMB\ unites functionality useful in many applied settings into a single package.
% TODO: mention this in response doc

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\textwidth]{figs/timing-spatial-multipanel-log.pdf}
\caption[]{
  \added{
  Same as Figure~\ref{fig:timing} but with both axes log distributed.
}
}
\label{fig:timing2}
\end{figure}

@@

% \input{../../../sdmTMB-paper-response/response2}

>>>>

x <- seq(0, 1, length.out = 200)
r <- seq(0.2, 1, 0.2)
r <- c(0.2, 0.6)
df <- data.frame(
  x = rep(x, length(r)),
  range = rep(r, each = length(x))
)
matern <- function(h, sigma = 1, kappa, nu = 1) {
  ret <- (sigma^2/(2^(nu - 1) * gamma(nu))) *
    ((kappa * abs(h))^nu) *
    besselK(kappa * abs(h), nu)
  ret
  ret[x == 0] <- sigma^2
  ret
}
blues <- RColorBrewer::brewer.pal(length(r) + 1, "Blues")[-1]
df$cor <- matern(df$x, kappa = sqrt(8) / df$range)
sim_g3 <- ggplot(df, aes(x, cor, col = as.factor(range), group = as.factor(range))) +
  geom_line() +
  theme_light() +
  xlab("Distance") +
  ylab("Correlation") +
  labs(colour = "Range") +
  coord_cartesian(expand = FALSE, ylim = c(0, 1)) +
  scale_colour_manual(values = blues) +
  geom_hline(yintercept = 0.13, col = "grey50", lty = 2) +
  scale_x_continuous(breaks = r) +
  theme(legend.position = "inside", legend.position.inside = c(0.7, 0.7)) +
  ggtitle("(c) Matérn correlation function")

@@

% We omit example code for Moran's I calculation for brevity but include it in the code supplement.
\added{
We see strong evidence for spatial autocorrelation for the model without a random field (p $<$ 0.01) but a lack of evidence for spatial correlation after including a random field suggesting that residual spatial autocorrelation is alleviated by including the random field.
The specific p-value is dependent on the seed due to the randomization in the randomized quantile residuals.
}

@@

\added{
where $n$ represents the number of left-out data points, $\log \mathcal{L}$ is a symbol for log likelihood, $y_i$ represents left-out data point $i$, and $\widehat{y_i}$ represents the prediction for left-out data point $i$ with the fixed effects at their MLEs and random effects at their empirical Bayes estimates.
}

@@

In practice, we would repeat this procedure several times to ensure the rank order is \added{insensitive} to the randomly chosen folds and, if it was, consider averaging across multiple folds or increasing the number of folds.

@@

<<dog-index, echo=TRUE>>=
grid$area <- 4 # 2 km x 2 km
pred2 <- predict(fit, newdata = grid, return_tmb_object = TRUE)
ind <- get_index(pred2, bias_correct = TRUE, area = grid$area)
@

@@

In our final example, we demonstrate a model with spatially varying coefficient (SVC) effects and illustrate combining uncertainty from parameters by working with draws from the joint parameter precision matrix.
SVC models are a class of models in which coefficients are allowed to vary spatially constrained by some smooth function \citep{hastie1993, thorson2023SVC}.
\deleted{In addition, we illustrate using restricted maximum likelihood (REML), which integrates over both the fixed and random effects.}

@@

snow_g1 <- ggplot(data = nao_effect_df) +
  geom_sf(data = land, fill = "white", colour = "white", lwd = 0.35) +
  geom_sf(data = lakes, colour = "gray23", fill = "grey90", lwd = 0.35) +
  geom_point(aes(X, Y, colour = nao_effect, size = count), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50", fill = NA, lwd = 0.35) +
  geom_sf(data = lakes, colour = "gray50", fill = NA, lwd = 0.35) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr)),
    guide = guide_colourbar(direction = "horizontal", title.vjust = 1, title.position = "top", label.position = "bottom")
  ) +
  guides(size = "none") +
  ggtitle(paste0("(a) Median multiplicative effect: ", round(min(snow$nao_effect), 2), " to ", round(max(snow$nao_effect), 2))) +
  labs(colour = "NAO effect\non Snowy Owl counts") +
  theme_bw() +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.25, 0.17),
    legend.title = element_text(size = 9, hjust = 0),
    plot.margin = unit(c(0, 0.5, 0, 0), "cm"),
    panel.background = element_rect(fill = "grey90", colour = NA),
    panel.grid.major = element_line(colour = "white", linewidth = 0.5),
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.box.background = element_rect(fill = "transparent", colour = NA),
    plot.title = element_text(size = 10, hjust = -0.15),
    axis.title = element_blank()
  )

th <- theme(
    legend.position = "none",
    panel.background = element_rect(fill = "grey90", colour = NA),
    panel.grid.major = element_line(colour = "white", linewidth = 0.5),
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.box.background = element_rect(fill = "transparent", colour = NA),
    plot.title = element_text(size = 10), axis.ticks = element_blank(),
    axis.title = element_blank(), axis.text = element_blank()
  )

snow_g2 <- ggplot(data = nao_effect_df) +
  geom_sf(data = land, fill = "white", colour = "white", lwd = 0.35) +
  geom_sf(data = lakes, colour = "gray23", fill = "grey90", lwd = 0.35) +
  geom_point(aes(X, Y, colour = nao_effect_lwr, size = count), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50", lwd = 0.35) +
  geom_sf(data = lakes, colour = "gray50", fill = NA, lwd = 0.35) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr))
  ) +
  ggtitle(paste0("(b) Lower 95% CI: ", round(min(snow$nao_effect_lwr), 2), " to ", round(max(snow$nao_effect_lwr), 2))) +
  theme_bw() +
  th

snow_g3 <- ggplot(data = nao_effect_df) +
  geom_sf(data = land, fill = "white", colour = "white", lwd = 0.35) +
  geom_sf(data = lakes, colour = "gray23", fill = "grey90", lwd = 0.35) +
  geom_point(aes(X, Y, colour = nao_effect_upr, size = count), alpha = 0.5) +
  geom_sf(data = coast, colour = "gray50", lwd = 0.35) +
  geom_sf(data = lakes, colour = "gray50", fill = NA, lwd = 0.35) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
  scale_colour_viridis_c(
    limit = c(min(snow$nao_effect_lwr), max(snow$nao_effect_upr))
  ) +
  ggtitle(paste0("(c) Upper 95% CI: ", round(min(snow$nao_effect_upr), 2), " to ", round(max(snow$nao_effect_upr), 2))) +
  theme_bw() +
  th

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\textwidth]{../../figs/timing-spatial-multipanel.pdf}
\caption[]{
Comparison of time to fit an SPDE spatial random field model with an intercept, one fixed-effect predictor, Gaussian error, and a sequence of SPDE resolutions to \added{three dataset sizes}.
Lines represent means and ribbons 90\% quantiles across 50 random iterations.
\pkg{VAST} and \pkg{tinyVAST} should be similar to \sdmTMB\ and so are not shown.
\pkg{inlabru} \added{(version 2.10.1)} used the empirical Bayes integration strategy and Gaussian approximation with \code{bru\_max\_iter = 1}, and the \fct{like} formulation.
\pkg{mgcv} \added{(version 1.9.1)} used \fct{bam}, \code{method = `fREML'}, and discretized covariates \citep{miller2019}.
Note that \pkg{spaMM} \added{(version 4.4.16)} only fits spatial, not spatiotemporal, models.
All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.
\added{See Figure~\ref{fig:timing2} for a version with log-log axes.}
}
\label{fig:timing}
\end{figure}

@@

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table~\ref{tab:functionality}), we suggest the primary differences are the user interface and speed.
We think users familiar with \fct{stats::glm}, \pkg{lme4}, \pkg{mgcv}, or \pkg{glmmTMB} will find \sdmTMB\ most approachable.
Users familiar with \INLA\ will find \pkg{inlabru} approachable.
Users familiar with \pkg{mgcv} can adapt \pkg{mgcv} to fit similar models with custom code \citep{miller2019} and \INLA/\pkg{inlabru} and \pkg{mgcv} are also general purpose modelling packages.
\pkg{VAST} and \pkg{tinyVAST} are the sole options for fitting some multivariate models; alternatively, because these packages focus on multivariate delta models and fisheries applications, users fitting ``simple'' univariate spatial/spatiotemporal GLMMs will likely find \sdmTMB\ more straightforward.
Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \sdmTMB\ or \pkg{VAST} (although such quantities can be calculated using other packages post hoc).
Overall, \deleted{we think there is value in that} \sdmTMB\ unites functionality useful in many applied settings into a single package.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\textwidth]{../../figs/timing-spatial-multipanel-log.pdf}
\caption[]{
  \added{
  Same as Figure~\ref{fig:timing} but with both axes log distributed.
}
}
\label{fig:timing2}
\end{figure}

@@

% \clearpage
% \input{../../../sdmTMB-paper-response/response2}
